---
title: "Viral diversity in fire ant queens"
authors: Anindita Brahma, Yannick Wurm
output: viral_diversity_fire_ants.html
---
#Introduction
This script contains comparison of viral diversity between fire ant queens from single- and multiple-queen colonies.
Input datasets include metadata (.csv) generated from counting the number of viral species present in each sample.
Metadata is available in supplementary datasets (Metadata for Fig 1B, Metadata for Fig 1D)

```{r setup, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
libraries <- c("dplyr", "ape", "geiger", "phytools",
               "nlme", "ggplot2", "lintr", "lme4", "MASS",
               "RColorBrewer", "viridis", "shades", "effects", "DHARMa")
for (lib in libraries) {
  if (require(package = lib, character.only = TRUE)) {
    message("Library loaded: ", lib)
  } else {
    print("Installing")
    BiocManager::install(lib, Ncpus = 10, update = TRUE)
    library(lib, character.only = TRUE)
    message("Library loaded: ", lib)
  }
}
```
#Analysis with all six datasets: Metadata for Fig 1B (supplementary dataset)

```{r}
#Import dataset
metadata <- read.table(file = "metadata.csv",
                                        header = TRUE, sep = ",")
#Fit the model
viral_diversity <- glmer(virus_species_count ~ social_form +
                                 (1|experiment_id),
                                 family = poisson(link = "log"),
                               data = metadata)
summary(viral_diversity)

#Check for overdispersion and non-uniformity of residuals
sim_output <- simulateResiduals(fittedModel = viral_diversity)
testDispersion(simulationOutput = sim_output)
testUniformity(sim_output)

#Calculate model estimates and confidence intervals
summary(allEffects(viral_diversity))

#Figure 1B
viral_diversity_boxplot <- ggplot(data = metadata,
                          mapping = aes(x = social_form,
                                        y = virus_species_count,
                                        fill = social_form)) +
  geom_boxplot() +
  geom_dotplot(binaxis = 'y', stackdir = 'center', binwidth = 0.15,
               stackratio = 1, position = position_dodge(1)) +
  theme_bw() +
  ylab("Number of viral species") +
  xlab("Social form") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_brewer(palette = "Greys") +
  scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10, 12))
```

#Focused analysis with one dataset: Metadata for Fig 1D (supplementary dataset)
```{r}
#Import dataset
metadata <- read.table(file = "metadata.csv",
                                        header = TRUE, sep = ",")
#Fit the model
viral_diversity_focused <- glmer(virus_species_count ~ social_form + genotype +
                           (1|colony_id/sample_id),
                         family = poisson(link = "log"),
                         data = metadata)
summary(viral_diversity_focused)

#Check for overdispersion and non-uniformity of residuals
sim_output <- simulateResiduals(fittedModel = viral_diversity_focused)
testDispersion(simulationOutput = sim_output)
testUniformity(sim_output)

#Calculate model estimates and confidence intervals
summary(allEffects(viral_diversity_focused))

#Figure 1D
viral_diversity_focused_boxplot <- ggplot(data = metadata,
                                          mapping = aes(x = social_form,
                                                        y = virus_species_count,
                                                        fill = genotype)) +
  geom_boxplot() +
  geom_dotplot(binaxis = 'y', stackdir = 'center', binwidth = 0.08,
               stackratio = 1, position = position_dodge(1)) +
  theme_bw() +
  ylab("Number of viral species") +
  xlab("Social form") +
  scale_y_continuous(breaks = c(0, 2, 4, 6, 8)) +
  scale_fill_brewer(palette = "Greys")

```