---
title: "viral_load_metadata"
authors: Anindita Brahma, Yannick Wurm
output: viral_load_metadata.html
---

#Introduction
This script is for preparing metadata for viral load analyses.
Input data contains extracted viral reads from Kraken2 reports.

```{r setup, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
libraries <- c("dplyr", "lme4",
               "ggplot2", "lintr", "plyr")
for (lib in libraries) {
  if (require(package = lib, character.only = TRUE)) {
    message("Library loaded: ", lib)
  } else {
    print("Installing")
    BiocManager::install(lib, Ncpus = 10, update = TRUE)
    library(lib, character.only = TRUE)
    message("Library loaded: ", lib)
  }
}

```
#Step 1. Prepare metadata for viral load analysis with all six datasets
#Extract viral families containing more than 100 reads.
```{r}
# Write a function
# sample: sample_id_viral_load.txt obtained from 02-Extract_virus_reads
# sample_details: Sample details for Fig 1A&B (supplementary dataset)

virus_reads <- function(sample, sample_details, id, social_form, tissue) {
   sample <- read.table(file = sample, header = FALSE,
                        sep = "\t", quote = "")
   read_data <- read.table(file = sample_details, header = TRUE,
                           sep = ",")
   colnames(sample) <- c("percentage_reads", "fragment_number",
                       "fragment_assigned", "tax_unit",
                       "taxid" ,"name")
   sample$id <- rep(x = id, times = length(sample$name))
   sample$social_form <- rep(x = social_form,
                             times = length(sample$name))
   sample$tissue <- rep(x = tissue,
                       times = length(sample$name))
   
   #Subset for family
   sample_family <- subset(x = sample, subset = tax_unit == "F")
   
   #Remove whitespace from name
   family_name <- data.frame(gsub('\\s+', '', sample_family$name))
   colnames(family_name) <- "family_name"
   
   #Add column
   sample_family$family_name <- family_name$family_name
   
   #Sample species with fragment number > 100
   sample_family_subset <- subset(x = sample_family,
                                   subset = fragment_number > 100)
   
   #Get caste, read counts, experiment_id
   sample_caste <- subset(x = read_data$caste,
                          subset = read_data$id == id)
   sample_family_subset$caste <- rep(x = sample_caste,
                              times = length(sample_family_subset$id))
   sample_experiment_id <- subset(x = read_data$experiment_id,
                          subset = read_data$id == id)
   sample_family_subset$experiment_id <- rep(x = sample_experiment_id,
                              times = length(sample_family_subset$id))
   
   total_read <- subset(x = read_data$read_ant_virus,
                        subset = read_data$id == id)
   sample_family_subset$total_read <- rep(x = total_read,
                                   times = length(sample_family_subset$id))
   
   #Normalise virus counts
   sample_family_subset$proportion_read <- sample_family_subset$fragment_number / sample_family_subset$total_read
   
   #Remove unnecessary columns
   sample_family_subset$name <- NULL
   sample_family_subset$percentage_reads <- NULL
   return(sample_family_subset)
}

# Use this function to extract virus families from sample_id_viral_load.txt
# Next combine all outputs into a single file: viral_load_metadata

# Write the combined file into a directory
write.table(x = viral_load_metadata, file = "viral_load_metadata.csv",
            sep = ",", col.names = TRUE, row.names = FALSE)

```
#Step 2. Prepare metadata for focused viral load analysis with one dataset.
#Extract viral families containing more than 100 reads.

```{r}
#Write a function
# sample: sample_id_viral_load.txt obtained from 02-Extract_virus_reads
# sample_details: Sample details for Fig 1C&D (supplementary dataset)

virus_reads_focused <- function(sample, sample_details, id, social_form, tissue, genotype, colony_id, sample_id) {
   sample <- read.table(file = sample, header = FALSE,
                        sep = "\t", quote = "")
   read_data <- read.table(file = sample_details, header = TRUE,
                           sep = ",")
   colnames(sample) <- c("percentage_reads", "fragment_number",
                       "fragment_assigned", "tax_unit",
                       "taxid" ,"name")
   sample$id <- rep(x = id, times = length(sample$name))
   sample$social_form <- rep(x = social_form,
                             times = length(sample$name))
   sample$tissue <- rep(x = tissue,
                       times = length(sample$name))
   sample$genotype <- rep(x = genotype,
                          times = length(sample$name))
   sample$colony_id <- rep(x = colony_id,
                          times = length(sample$name))
   sample$sample_id <- rep(x = sample_id,
                          times = length(sample$name))
   #Subset for family
   sample_family <- subset(x = sample, subset = tax_unit == "F")
   
   #Remove whitespace from name
   family_name <- data.frame(gsub('\\s+', '', sample_family$name))
   colnames(family_name) <- "family_name"
   
   #Add column
   sample_family$family_name <- family_name$family_name
   
   #Sample families with fragment number > 100
   sample_family_subset <- subset(x = sample_family,
                                   subset = fragment_number > 100)
   #Get caste, read counts, experiment_id
   sample_caste <- subset(x = read_data$caste,
                          subset = read_data$id == id)
   sample_family_subset$caste <- rep(x = sample_caste,
                              times = length(sample_family_subset$id))
   sample_experiment_id <- subset(x = read_data$experiment_id,
                          subset = read_data$id == id)
   sample_family_subset$experiment_id <- rep(x = sample_experiment_id,
                              times = length(sample_family_subset$id))
   
   total_read <- subset(x = read_data$read_ant_virus,
                        subset = read_data$id == id)
   sample_family_subset$total_read <- rep(x = total_read,
                                   times = length(sample_family_subset$id))
   #Normalise virus counts
   sample_family_subset$proportion_read <- sample_family_subset$fragment_number / sample_family_subset$total_read
   
   #Remove unnecessary columns
   sample_family_subset$name <- NULL
   sample_family_subset$percentage_reads <- NULL
   return(sample_family_subset)
}

# Use this function to extract virus families from sample_id_viral_load.txt
# Next combine all outputs into a single file: viral_load_focused_metadata

# Write the combined file into a directory
write.table(x = viral_load_focused_metadata, file = "viral_load_focused_metadata.csv",
            sep = ",", col.names = TRUE, row.names = FALSE)

```
