---
title: "Viral load in fire ant queens"
authors: Anindita Brahma, Yannick Wurm
output: viral_load_fire_ants.html
---
#Introduction
This script contains comparison of viral load between fire ant queens from single- and multiple-queen colonies.
Input datasets include metadata (.csv) generated from extracted reads assigned to viral families.
Metadata is available in supplementary dataset (Metadata for Fig 1A, Metadata for Fig 1C)

```{r setup, include=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
libraries <- c("dplyr", "lme4", "MASS", "lattice",
               "glmmTMB", "car", "dplyr", "blme",
               "logspline", "fitdistrplus",
               "ggplot2", "ggpubr", "lintr",
               "RColorBrewer", "scales", "DHARMa", "effects")
for (lib in libraries) {
  if (require(package = lib, character.only = TRUE)) {
    message("Library loaded: ", lib)
  } else {
    print("Installing")
    BiocManager::install(lib, Ncpus = 10, update = TRUE)
    library(lib, character.only = TRUE)
    message("Library loaded: ", lib)
  }
}

```
#Analysis with all six datasets: Metadata for Fig 1A (supplementary dataset)

```{r}
#Import dataset
metadata <- read.table(file = "metadata.csv",
                                        header = TRUE, sep = ",")

#Add a column for percentage_read for each viral family
metadata$percentage_read <- (metadata$proportion_read) * 100

#Calculate sum of proportion read for each individual
all_id_vec <- unique(metadata$sample_id)

#Create dataframe to gather information from each unique id
all_id_sum_of_reads <- as.data.frame(matrix(ncol = 4,
                                            nrow = length(all_id_vec)))

colnames(all_id_sum_of_reads) <- c("sample_id", "total_virus_reads", "sum_proportion_read", "sum_percentage_read")

all_id_sum_of_reads$id <- all_id_vec

#Gather details for each sample
sample_details <- metadata %>% distinct(sample_id, social_form, tissue, caste,
                                                  experiment_id, total_read)

#Add the proportion_reads and percentage_reads from all viral famililies for each sample
for (position in seq_len(length(all_id_vec))) {
   print(position)
   
  #Subset metadata for each sample
   virus_read <- subset(x = metadata,
                              subset = id == all_id_vec[position],
                              select = fragment_number)
   prop_read <- subset(x = metadata,
                       subset = id == all_id_vec[position],
                       select = proportion_read)
   percent_read <- subset(x = metadata,
                       subset = id == all_id_vec[position],
                       select = percentage_read)
                       
   #Sum of toral_virus_reads, proportion read, percentage_reads
   total_virus_read <- sum(data.matrix(virus_read))
   sum_prop_read <- sum(data.matrix(prop_read))
   sum_percent_read <- sum(data.matrix(percent_read))
   all_id_sum_of_reads$total_virus_reads[position] <- total_virus_read
   all_id_sum_of_reads$sum_proportion_read[position] <- sum_prop_read
   all_id_sum_of_reads$sum_percentage_read[position] <- sum_percent_read
   
}

#Add the sample_details
all_id_prop_read <- merge(x = sample_details,
                                      y = all_id_sum_of_reads)
#Fit the model
viral_load <- glmer(sum_proportion_read ~ social_form +
                                      (1|experiment_id/sample_id),
                               family = binomial(link = "log"),
                               weights = total_read,
                               data = all_id_prop_read)
                  
summary(viral_load)

#Check for overdispersion and non-uniformity of residuals
sim_output <- simulateResiduals(fittedModel = viral_load)
testDispersion(simulationOutput = sim_output)
testUniformity(sim_output)

#Calculate model estimates and confidence intervals
summary(allEffects(viral_load))

#Figure 1A
viral_load_boxplot <- ggplot(data = all_id_prop_read,
       mapping = aes(x = social_form, y = sum_percentage_read, fill = social_form)) +
  geom_boxplot() +
  geom_dotplot(binaxis = 'y', stackdir = 'center', binwidth = 0.06,
               stackratio = 1, position = position_dodge(1)) +
  ylab("Percentage of viral reads") +
  xlab("Social form") +
  theme_bw() +
  scale_y_continuous(trans = "log", breaks = c(0.01, 1, 20)) +
  scale_fill_manual(values = palette_metadata)


```
#Focused analysis with one dataset: Metadata for Fig 1C (supplementary dataset)
```{r}
#Import dataset
metadata <- read.table(file = "metadata",
                                   header = TRUE, sep = ",")
#Add a column for percentage_read for each viral family
metadata$percentage_read <- (metadata$proportion_read) * 100

#Calculate sum of proportion read for each individual
all_id_vec <- unique(metadata$sample_id)

#Create dataframe to gather information from each unique id
all_id_sum_of_reads <- as.data.frame(matrix(ncol = 4,
                                            nrow = length(all_id_vec)))

colnames(all_id_sum_of_reads) <- c("sample_id", "total_virus_reads", "sum_proportion_read", "sum_percentage_read")

all_id_sum_of_reads$id <- all_id_vec

#Gather details for each sample
sample_details <- metadata %>% distinct(sample_id, social_form, caste, colony_id,
                                                 genotype, experiment_id, total_read)

#Add the proportion_reads and percentage_reads from all viral famililies for each sample
for (position in seq_len(length(all_id_vec))) {
   print(position)
   
  #Subset metadata for each sample
   virus_read <- subset(x = metadata,
                              subset = id == all_id_vec[position],
                              select = fragment_number)
   prop_read <- subset(x = metadata,
                       subset = id == all_id_vec[position],
                       select = proportion_read)
   percent_read <- subset(x = metadata,
                       subset = id == all_id_vec[position],
                       select = percentage_read)
                       
   #Sum of toral_virus_reads, proportion read, percentage_reads
   total_virus_read <- sum(data.matrix(virus_read))
   sum_prop_read <- sum(data.matrix(prop_read))
   sum_percent_read <- sum(data.matrix(percent_read))
   all_id_sum_of_reads$total_virus_reads[position] <- total_virus_read
   all_id_sum_of_reads$sum_proportion_read[position] <- sum_prop_read
   all_id_sum_of_reads$sum_percentage_read[position] <- sum_percent_read
   
}

#Add the sample_details
all_id_prop_read <- merge(x = sample_details,
                                      y = all_id_sum_of_reads)

#Fit the model
viral_load_focused <- glmer(sum_proportion_read ~ social_form + genotype +
                                      (1|colony_id/sample_id),
                               family = binomial(link = "log"),
                               weights = total_read,
                               data = all_id_prop_read)
                  
summary(viral_load_focused)

#Check for overdispersion and non-uniformity of residuals
sim_output <- simulateResiduals(fittedModel = viral_load_focused)
testDispersion(simulationOutput = sim_output)
testUniformity(sim_output)

#Calculate model estimates and confidence intervals
summary(allEffects(viral_load_focused))

#Figure 1C
viral_load_focused_boxplot <- ggplot(data = all_id_prop_read,
       mapping = aes(x = social_form, y = sum_percentage_read, fill = genotype)) +
  geom_boxplot() +
  geom_dotplot(binaxis = 'y', stackdir = 'center', binwidth = 0.06,
               stackratio = 1, position = position_dodge(1)) +
  ylab("Percentage of viral reads") +
  xlab("Social form") +
  theme_bw() +
   scale_y_continuous(trans = "log", breaks = c(0.01, 1, 10)) +
   scale_fill_brewer(palette = "Greys")

```
